Precision calculations for cosmic shear

### 2017-02-08 discussion.tex
MK cleaned up appendix B

The two-point shear correlation function $\xi_\pm$ represents the current baseline observable for cosmic shear measurements.   
As shown in Figure~\ref{fig:xi_pm_K16}, however, using the standard first-order extended Limber flat-sky approximation (ExtL1FlHyb) can result
in errors exceeding 10 percent, on angular scales $\theta > 300$ arcmin.    This is a result of the significant weight given to low $\ell$ modes in the $\xi_+$ statistic, as illustrated in Figure~\ref{fig:filters} which shows the integrand of $\xi_+$ and $\xi_-$ (upper two panels) for two cases ( $\theta = 100$ and $\theta = 350$ arcmin), normalised to their maximum
value.   This error does not impact CFHTLenS analyses, given the low signal-to-noise of the measurements on these scales.  It will however become increasingly important for upcoming wider-field surveys which will accurately probe these scales.

>>>>

The two-point shear correlation function $\xi_\pm$ represents the current
baseline observable for cosmic shear measurements.   As shown in
Figure~\ref{fig:xi_pm}, however, using the standard first-order extended
Limber flat-sky approximation (ExtL1FlHyb) can result in errors exceeding 10
percent, on angular scales $\theta > 300$ arcmin.    This is a result of the
significant weight given to low $\ell$ modes in the $\xi_+$ statistic, as
illustrated in Figure~\ref{fig:filters} which shows the integrand of $\xi_+$
and $\xi_-$ (upper two panels) for two cases ( $\theta = 100$ and $\theta =
350$ arcmin), normalised to their maximum value.   This error does not impact
CFHTLenS analyses, given the low signal-to-noise of the measurements on these
scales.  It will however become increasingly important for upcoming wider-field
surveys which will accurately probe these scales.

### 2017-02-08 limber2.tex
MK cleaned up appendix B

Figure.~\ref{fig:xi_pm_K16} presents the two-point correlation functions
$\xi_+$ (left) and $\xi_-$ (right) using the different cases for the shear
power spectrum listed in Table~\ref{tab:cases}. The adopted CFHTLenS redshift
distribution and fiducial cosmological model are the same as in
Figure.~\ref{fig:Cl_cases}. As is clear by the red dotted curve, using the
Planck cosmology \citep{2015arXiv150201589P} induces a much larger change in
the amplitude of the shear correlation function than the different projection
methods.

@@

  \label{fig:xi_pm_K16}

@@

\subsection{Kitching et al.~2016}

\cite{2016arXiv161104954K} compute the full projection of the weak-lensing power
spectrum, which they present as spherical-radial representation the 3D shear field.
Our results (\ref{eq:C_ell_full}) correspond to their equations (7) and (8)
for a flat Universe and in the case of perfect
photometric redshifts, $p(z | z_{\rm p}) = \delta_{\rm D}(z - z_{\rm p})$, and
for a bin function that selects the redshift bin $i$, $W^{\rm SR}(z, z_{\rm p})
= 1_{z \in i-\mbox{th bin}}$. In \cite{2016arXiv161104954K} eq.~(7) is however the
factor $2/\pi$ missing.

In App.~A \cite{2016arXiv161104954K} derives the spherical and extended
Limber approximation starting from the full spherical projection. Note that
their filter function $q$ defined in (31) has an errornous factor of comoving
distance $r$, and an additional factor $\pi/2$ appears. \mk{This is probably worded
too strongly.}

We can not reproduce the amplitude of the differences between the cases,
neither for the power spectrum nor for the shear correlation function (see
Figs.~\ref{fig:Cl_cases} and \ref{fig:xi_pm_K16}).

\paragraph{Bernardeau, Bonvin, Van de Rijt, Vernizzi (2012); van de Rijt (2012)}

Eq.~(44) of \cite{2012PhRvD..86b3001B} is the non-tomographic $C(\ell)$ in
terms of the time-independent potential power spectrum $T^2(k) P(k)$, where $T$
is the transfer function; the growth factor $D$ is contained in their window
function $W$ (43) as growth-suppression factor $g(\chi) = D(\chi)/a(\chi)$
(defined after eq.~{24}). For a single source redshift, we find the identity
$W(\chi)$ = $q(\chi)\chi^{-1} D(\chi) a^{-1}(\chi)$.
\cite{2012PhRvD..86b3001B} is consistent with our
(\ref{eq:C_ell_full}), if we interpret the
scale factors at distances $\chi$ and $\chi^\prime$, respectively.

The PhD thesis of Van de Rijt \cite{vande2012} presents an explicit calculation
of the second-order Limber approximation for a fixed source redshift. They
carry out the derivatives of the kernels $f$ under the assumption of a constant
growth-suppression factor $D_+(a)/a$, and a power spectrum that separates $k$-
from $\chi$-terms. We show the consistency between their result (eq.~7.19) and
our (\ref{eq:C_ell_limber2_dr_s}), as follows.

We write (\ref{eq:C_ell_limber2_dr_s}) without tomography, and insert again the
time-independent power spectrum (\ref{eq:P_geom_mean}),
%
\begin{align}
  C^\gamma_{{\rm L2}}(\ell) = & - \frac 1 {\nu^2} \, \frac{\ellbar^2(\ell, 2)}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi^{2} P_{\rm m}\left(k = \frac\nu\chi\right)
    \nonumber \\
    & \times D_+(\chi) a^{-1}(\chi) \chi^{3/2} q(\chi) \left[ f^{\prime\prime} 
      + \frac{\chi}{3} f^{\prime\prime\prime} 
    \right](\chi)
  \label{eq:C_ell_limber2_dr_notomo}
\end{align}
%
For a fixed source comoving distance $\chi_{\rm S}$ the lensing efficiency is
$q(\chi) = (\chi_{\rm S} - \chi) \chi_{\rm S}^{-1}$. Assuming that
$D_+(\chi)/a(\chi) \approx g = \mbox{const}$, we get for the derivative terms,
with $f(\chi) = g \chi_{\rm S}^{-1} (\chi_{\rm S} - \chi) \chi^{-3/2}$,
%
\begin{equation}
  f^{\prime\prime}(\chi) + \frac{\chi}{3} f^{\prime\prime\prime}(\chi)
    = - \frac 1 8 \chi^{-5/2} \left( \frac 5 \chi
          + \frac 1 {\chi_{\rm S}} \right) \frac{D_+(\chi)}{a(\chi)}.
\end{equation}
%
Therefore,
%
\begin{align}
  C^\gamma_{{\rm L2}}(\ell) = \frac 1 {\nu^2} \, \frac{\ellbar^2(\ell, 2)}{\nu^4} \, \pref^2
    \int {\rm d} \chi \, \chi \, P_{\rm m}\left(k = \frac\nu\chi\right)
    \frac{q(\chi)}{8} \left( \frac 5 \chi + \frac 1 {\chi_{\rm S}} \right)
      \left(\frac{D_+(\chi)}{a(\chi)}\right)^2 
\end{align}
%
We now replace the matter with the potential power spectrum, ${\cal A}^2 P_{\rm m}(k) = k^{4} a^2 P_\Phi(k)$, and pull into the integral $\nu^{-4} = \chi^{-4} k^{-4}$. This yields
%
\begin{align}
  C^\gamma_{{\rm L2}}(\ell) = \frac 1 {\nu^2} \, \ellbar^2(\ell, 2)
    \int \frac{{\rm d} \chi}{\chi^2} \, P_\Phi\left(k = \frac\nu\chi\right)
    a^2(\chi) \frac{1}{8} \frac{\chi_{\rm S} - \chi}{\chi \chi_{\rm S}}
    \left( \frac 5 \chi + \frac 1 {\chi_{\rm S}} \right)
      \left(\frac{D_+(\chi)}{a(\chi)}\right)^2 .
\end{align}
%
Note the additional factor $a^2(\chi)$. This difference is accounted for by the
factor $a^{-1}$ in the propagation of primordial potential perturbations to
late times, see eq.~(6.22) in \cite{vande2012}, where the transition is done
using $g(a)$ instead of $D(a)$. \mk{Check this further.}

@@

\paragraph{LoVerde \& Afshordi (2009)}

This paper introduces the extended Limber approximation to second order that we
apply in this work. Although they present the specific case of 2D galaxy
clustering, their calculations are general enough to apply to weak lensing.

Their eq.~(5) is a general spherical power spectrum of a scalar field,
projected from 3D to 2D via (4). Comparing their expressions with the
weak-lensing potential (\ref{eq:psi}), we can set $F_A(\chi) = F_B(\chi) = 2
c^{-2} D_+(\chi) q(\chi) \chi^{-1}$. With that, their eq.~(5) is identical to
our (\ref{eq:C_ell_phi_Pphi}).

The second-order Limber approximation in \cite{2008PhRvD..78l3506L} is presented in
eq.~(12). This is consistent with our first-order (\ref{eq:C_ell_limber1}) and
second-order (\ref{eq:C_ell_limber2_dr}) Limber approximation, when accounting for the difference between
lensing potential and shear 2D power spectrum, and 3D potential and matter power spectrum.

\paragraph{Schmidt (2009)}

In eq.~(9) \cite{2008PhRvD..78d3002S} write the lensing power spectrum in the
flat-sky limit. The Poisson factor $D_\phi$ is defined in their eq.~(5), see
also our eq.~\ref{eq:poisson}, and $D_{\rm m}$ is the growth factor. The filter
function $W_\kappa$ (eq.~10) can be written as
%
\begin{align}
  W_\kappa(z) = & \frac 1 {H(z)} \int_z^\infty {\rm d} z_{\rm s} W_{\rm L}(z_{\rm s}, z) W_{\rm g}(z)
              =  \frac 1 {H(z)} \chi \int_\chi^\infty {\rm d} \chi^\prime \frac{\chi^\prime - \chi}{\chi^\prime}
                n(\chi)
              \nonumber \\
              & \equiv \frac 1 {H(z)} \chi q(\chi),
\end{align}
%
using ${\rm d} z W_{\rm g}(z) = {\rm d} \chi n(\chi)$.
To write the redshift integrals (9) in \cite{2008PhRvD..78d3002S} over 
comoving distance, we make the variable transformation
%
\begin{equation}
  {\rm d} \chi = \frac{c {\rm d} z}{H(z)} .
\end{equation}
%
We recover our (\ref{eq:P_ell_gamma}).

\paragraph{Giannantonio et al. (2012)}

In eqs.~(25) and (26) \cite{2012MNRAS.422.2854G} write the flat-sky lensing
power spectrum. Their window function $W^{\varepsilon_i}$ defined in (25) for a
flat Universe equals $\pref \times q(\chi) / a(\chi)$, since ${\rm d} z {\rm d}
N(z)/{\rm d} z = {\rm d} \chi n(\chi)$; however there is a factor $r_K[r(z)]$
missing in the window function, which translates into a missing $r_K[r]
r_K[r^\prime]$ in the integral\footnote{Confirmed by T.~G., priv.~comm.}. This
also leads to an errorous $r_K^{-2}(r)$ in the Limber equation (27). With these
factors accounted for, we reproduce the expressions of \cite{2012MNRAS.422.2854G}.

>>>>

Figure.~\ref{fig:xi_pm} presents the two-point correlation functions
$\xi_+$ (left) and $\xi_-$ (right) using the different cases for the shear
power spectrum listed in Table~\ref{tab:cases}. The adopted CFHTLenS redshift
distribution and fiducial cosmological model are the same as in
Figure.~\ref{fig:Cl_cases}. As is clear by the red dotted curve, using the
Planck cosmology \citep{2015arXiv150201589P} induces a much larger change in
the amplitude of the shear correlation function than the different projection
methods.

@@

  \label{fig:xi_pm}

@@

In this section we briefly discuss previously published work on the full
projection and second-order Limber equation of the lensing power spectrum. We
compare those results with our findings.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Kitching et al.~2016}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\cite{2016arXiv161104954K} compute the full projection of the weak-lensing
power spectrum, which they present as spherical-radial representation the 3D
shear field. Our results (\ref{eq:C_ell_full}) correspond to their equations
(7) and (8) for a flat Universe and in the case of perfect photometric
redshifts, $p(z | z_{\rm p}) = \delta_{\rm D}(z - z_{\rm p})$, and for a bin
function that selects the redshift bin $i$, $W^{\rm SR}(z, z_{\rm p}) = 1_{z
\in i-\mbox{th bin}}$. In \cite{2016arXiv161104954K} eq.~(7) the factor $2/\pi$
is absent.

In App.~A \cite{2016arXiv161104954K} derives the spherical and extended Limber
approximation starting from the full spherical projection. Note that their
filter function $q$ defined in (31) has an errornous additional factor of
comoving distance $r$, and shows an additional factor of $\pi/2$.

As shown in the main text of our paper,
we can not reproduce the amplitude of the differences between the cases,
neither for the power spectrum nor for the shear correlation function (see
Figs.~\ref{fig:Cl_cases} and \ref{fig:xi_pm}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Bernardeau, Bonvin, Van de Rijt, Vernizzi (2012); van de Rijt (2012)}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Eq.~(44) of \cite{2012PhRvD..86b3001B} is the non-tomographic full projection
$C(\ell)$ in the approximation of the 3D potential power spectrum $P^\Phi$ separating
into $k$- and $\chi$-dependent functions. Their expression holds for a single
source redshift.

The PhD thesis of Van de Rijt \cite{vande2012} presents an explicit calculation
of the second-order Limber approximation. They carry out the derivatives of the
kernels $f$ under the assumption of a constant growth-suppression factor
$D_+(a)/a$. Then, for a constant source comoving distance $\chi_{\rm S}$, the
lensing efficiency (\ref{eq:lens_efficiency}) is $q(\chi) = (\chi_{\rm S} -
\chi) \chi_{\rm S}^{-1}$, and the derivative of the separated kernel function
(\ref{eq:f_LA08_s}) can be calculated analytically,
%
\begin{equation}
  f^{\prime\prime}(\chi) + \frac{\chi}{3} f^{\prime\prime\prime}(\chi)
    = - \frac 1 8 \chi^{-5/2} \left( \frac 5 \chi
          + \frac 1 {\chi_{\rm S}} \right) \frac{D_+(\chi)}{a(\chi)}.
\end{equation}
%
This result confirms the power-law behaviour of the filter function, which we exploited earlier to fit
this function.

Inserting this into the second-order Limber power spectrum
(\ref{eq:C_ell_limber2_dr_s}) and using the inverse Poisson equation to replace
the matter with the potential power spectrum we obtain the same expression as
\cite{vande2012} (their eq.~7.19), except for an additional factor $a^2(\chi)$.
This difference is accounted for by the factor $a^{-1}$ in the propagation of
primordial potential perturbations to late times, see eq.~(6.22) in
\cite{vande2012}, where the transition is done using $g(a)$ instead of $D(a)$.
\mk{Check this further.}

@@

\subsection{LoVerde \& Afshordi (2009)}

This paper introduces the extended Limber approximation to second order that we
apply in this work. Although they present the specific case of 2D galaxy
clustering, their calculations are general enough to apply to a weak-lensing
context. Their eq.~(5) is a spherical cross-power spectrum of two scalar fields
$A$ and $B$, projected from 3D to 2D via projection kernels $F_A$ and $F_B$
defined in their eq.~(4). Comparing their expressions with the weak-lensing
potential (\ref{eq:psi}), we set $F_A(\chi) = F_B(\chi) = 2 c^{-2}
D_+(\chi) q(\chi) \chi^{-1}$. With that, their eq.~(5) is identical to our
(\ref{eq:C_ell_phi_Pphi}).

The second-order Limber approximation in \cite{2008PhRvD..78l3506L} is
presented in eq.~(12). This is consistent with our first-order
(\ref{eq:C_ell_limber1}) and second-order (\ref{eq:C_ell_limber2_dr}) Limber
approximation terms, when accounting for the difference between lensing
potential and shear 2D power spectrum, and 3D potential and matter power
spectrum.

\subsection{Schmidt (2009)}

In eq.~(9) \cite{2008PhRvD..78d3002S} write the lensing power spectrum in the
flat-sky limit. Inserting the Poisson factor $D_\phi$, 
 the growth function $D_{\rm m}$, and writing the redshift filter
function $W_\kappa$ (eq.~10) in terms of the lensing efficiency and comoving distances,
$W_\kappa[z(\chi)] = H(z)^{-1} \chi q(\chi)$, we
%
%\begin{align}
  %W_\kappa(z) = & \frac 1 {H(z)} \int_z^\infty {\rm d} z_{\rm s} W_{\rm L}(z_{\rm s}, z) W_{\rm g}(z)
              %=  \frac 1 {H(z)} \chi \int_\chi^\infty {\rm d} \chi^\prime \frac{\chi^\prime - \chi}{\chi^\prime}
                %n(\chi)
              %\nonumber \\
              %& \equiv \frac 1 {H(z)} \chi q(\chi),
%\end{align}
%%
%using ${\rm d} z W_{\rm g}(z) = {\rm d} \chi n(\chi)$.
%To write the redshift integrals (9) in \cite{2008PhRvD..78d3002S} over 
%comoving distance, we make the variable transformation
%%
%\begin{equation}
  %{\rm d} \chi = \frac{c {\rm d} z}{H(z)} .
%\end{equation}
%%
%We
recover our (\ref{eq:P_ell_gamma}).

\subsection{Giannantonio et al. (2012)}

In eqs.~(25) and (26) \cite{2012MNRAS.422.2854G} write the flat-sky lensing
power spectrum. Their window function $W^{\varepsilon_i}$ defined in (25) for a
flat Universe equals $\pref \times q(\chi) / a(\chi)$, since ${\rm d} z \,
({\rm d} N(z)/{\rm d} z) = {\rm d} \chi n(\chi)$; however due to a typo there
is a factor $r_K[r(z)]$ missing in the window function, which translates into a
missing $r_K[r] r_K[r^\prime]$ in the full projection integral\footnote{Typo confirmed by T.~G.,
priv.~comm.}. This also leads to an errorous $r_K^{-2}(r)$ in the Limber
equation (27). With these factors accounted for, we reproduce the expressions
of \cite{2012MNRAS.422.2854G}.
